The tracing of trade

### 2021-02-11 Manuscript/Manuscript.tex
second review ready

To achieve this goal, we analyse a dataset of stamps found in two types of sites: 1) workshops from \textit{Baetica} province and 2) consumption centres located in two provinces \textit{Germania} (Inferior and Superior) and \textit{Britannia}. This dataset is used to detect links between specific sites that may share a large percentage of common stamps. The analysis is performed using quantitative methods borrowed from Ecology that allows us to group sites based on the number of stamps shared between them.

@@

The finding of identical amphoric stamps in different provinces has allowed us to explore patterns of olive oil distribution within the Empire. Despite this archaeological evidence, there is still no consensus on the structure that generated this pattern; how was the economic relation between production and consumption centres? Did they follow a centralised process to distribute olive oil in those provinces or each workshop organised its own trade networks?
%repite aims dos veces  
This paper aims to explore these questions regarding the olive oil market by identifying links between production centres and consumption provinces using similarity indexes for amphoric stamps. In this way, our work aims to detect regional patterns for a specific commercial product (i.e. olive oil) which was shipped between distant provinces~\citep{isaksen_network_2006}.

%We want to understand the pattern of olive oil production linked to amphora workshops and stamps used to mark them. 
%We focus here on exploring the economic relation between stamps and amphora production and distribution centres. 
This relation between the production centre (i.e. \textit{Baetica}) and consumption sites have been studied here for two case studies: \textit{Britannia} and \textit{Germania} (see Figure~\ref{general}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{general_map}
\caption{Sites analysed in this paper. The colour defines the three different regions under study (\textit{Baetica} (yellow) as producer; \textit{Germania} (blue) and \textit{Britannia} (green) as consumers) while the size of each dot is proportional to the number of Dressel 20 amphoric stamps found on each site}

@@

In \textit{Baetica} province we want to identify links between workshops based on the similarity of their stamps. In the consumption Roman provinces, our aim is to detect groups of nearby sites sharing stamps and to what extent stamps are clustered based on provinces. This economic connection could be identified with two ideas: a) nearby sites have more similar amphoric stamps and b) sets of stamps can only be found on specific provinces because they were shipped together.
%find in or find on

@@

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by identifying which centres have more similar stamps and then confront this measure against their proximity. To do this, a population-based approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. This quantitative method provides a way to measure the similarity between centre based on the amphoric stamps found on each of them. This ecological approach is based on three steps: a) to quantify stamp similarity between sites, b) to group sites based on this similarity and c) to find correlation patterns between stamp similarity and proximity. The rationale behind the method is as follows: if trade structure was centrally planned then we should be able to establish links between workshops and consumption sites based on the similarity of its stamps; if this is not the case then results would suggest that workshops worked independently and agreements were made at an individual level.

The paper addresses these questions, the analysis and results as follows: next section introduces the historical context. Section two defines the dataset and the methods used for the analysis. Section three presents the results and section four discusses them; the paper finishes with some concluding remarks on the research performed.

@@

The \textit{Baetican} province was one of the most suitable regions to tackle the challenge of the huge demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive specialised infrastructure of olive oil production was gradually deployed. This production of olive oil had the largest peak from the first to the third century AD~\citep{remesal_concierto}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{baetica}
\caption{Distribution of stamps found in Dressel 20 workshops along the Guadalquivir riverbank. The size of each point depends on the number of stamps found on each site}

@@

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were typically marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure} 
%found in find in
Most stamps seemed to have been used across a long activity period which can be difficult to date with the required level of accuracy. They have been frequently dated using as reference consular dates found on nearby \textit{tituli picti} recovered from \textit{Monte Testaccio}~\citep{Testaccio1, berni_millet_epigrafianforica_2008}.
The chronological information of \textit{tituli picti} is very useful, but the small amount of dates recovered and the fact that a vast majority of them come from the same site poses some challenges to this precise dating. In addition, the fact that \textit{Monte Testaccio} was a dump suggests that the disposition of amphorae vessels may have been altered at a later phase due to changes in the long-term dynamics of creation of this artificial mound.

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they could also be imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps was not defined by a single criteria, but they often displayed a code composed by letters and additional symbols. One of the most common forms is a code of three letters showing a \textit{Tria Nomina}~\citep{berni_millet_amphora_1996}; in this case the three letters are the initials of a Roman name. In some cases the full name may appear, while the meaning of other stamps is difficult to interpret~\citep{rodriguez_baetican_1998}.

@@

The expansion of the Roman Empire to increasingly distant regions required an effort to intensify trade activity through Mediterranean and Atlantic routes. These needs effectively led to a gradual change in the economic and social structures as the trade networks that supplied the provinces and particularly the Roman army became more complex. An interesting example of this system is the figure of the \textit{praefectura annonae} which was created during Augustus's administration to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced in the region as the landscape and climate of the British Isles was not suitable for the growth of olive trees~\citep[161]{monfort_britanniaen_1998}. Archaeological records suggest that this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D. onwards \citep{peacock_amphorae_1991,carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exports linked to the arrival of legions caused by different military campaigns~\citep[161]{monfort_britanniaen_1998}.

The increase of Dressel 20 exports created an important commercial network for exchanges. The demand for olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply. This economic relation will be particularly intense for sites close to Hadrian Wall's garrisons. Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic locations~\citep{carreras_atlantic_2012}.  This presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,funari_corpus_1996,carreras_britannia_1998,carreras_abastecimiento_2003}. This intense consumption of olive oil also suggests the existance of provincial structures created to organise the supply of olive oil across military camps. There are no written records detailing how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep{funari_economic_2005,orengo_seeds_2016}.

The consumption of Baetican olive oil would experience a progressive slowdown from the third century A.D. onwards. This date matches a change in market strategy in the Empire. This gradual decrease on \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by other types of amphorae from other places and particularly the African provinces~\citep{rodriguez1991aceite,millet_anforas_1998}.

@@

The distribution of Dressel 20 amphorae in the provinces of Germania Inferior and Germania Superior is not as well interpreted as the rest of provinces discussed in this work~\citep[293]{remesal_baetica_2002}. It is clear that there was a massive import of Dressel 20 in both provinces as is observed in archaeological excavations across the German \textit{limes}. However, it is still unknown to what extent trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}.

In any case, the introduction of olive oil in \textit{Germania} shows similar patterns than the one discussed for \textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region, while a majority of recovered Dressel 20 amphorae are located in the military sites built over the German \textit{limes} \citep{remesal_germaniaengl_2002}. The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a major concentration at the German \textit{limes}. It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons~\citep[125]{remesal_concierto}. However, this hypothesis is hard to assess given the current lack of written evidence clarifying the meaning of the amphoric stamps.

Some studies assumed that Roman naval technology was not suited for oceanic transport and for this reason defined riverine shipping across Rhône and Rhine rivers as the basic supply route for the German \textit{limes}. However, more recent large-scale computational analysis as well as new excavations on the Iberian Atlantic coast suggest that the Atlantic route, which was far more efficient, may have played a major role on the supply of \textit{Germania} and \textit{Britannia}~\citep{rubio-campillo_ecology_2018}.

@@

The goal of this study is to explore these economic and long-range trade dynamics using amphoric stamps as proxies for common supply networks. We are especially interested on identifying links between Baetican production centres and consumption centres in the different \textit{limes}.

For this analysis we used data collected in the CEIPAC database of amphoric epigraphy~\citep{remesal_centro_2015}. The CEIPAC dataset contains over 50.000 epigraphic records found on different types of amphorae (the database is openly accessible from \url{http://romanopendata.eu}). A large percentage of these records come from from \textit{Monte Testaccio}, but the rest is distributed mainly across the Western provinces of the Empire. It is worth mentioning that the database does not provide temporal information due to the reasons previously discussed, and for this reason this work can only analyze the spatial distributions within a coarse chronology of 2 centuries (3rd-1st AD).

@@

It is worth mentioning that some workshops were coded under the same geographical coordinates. For instance, the stamps of two nearby amphora workshops such as Tesorillo de Do\~na Mencia and Do\~na Mencia were catalogued under different names while sharing the same geographical coordinates. The interpretation of our analysis considered this bias in the archaeological dataset which may have a minor effect in the final results.

The first step of this analysis was to compute the frequency distribution of stamp codes and analyse it using Exploratory Data Analysis (EDA). This would allow us to study the distribution of the stamps across centres as can be seen in Figure~\ref{stamps}. As can be seen the distribution is widely diverse with most workshops having only one stamp, while a few workshops contained over 100 stamps. Particularly, one workshop (La Catria) concentrated a large percentage of the amphoric stamps with a total number of 228 stamps.

This uneven distribution of stamps is typical for amphora production as it is observed that a major concentration on the number of stamps is focused on a few workshops~\citep{coto-sarmiento_identifying_2018}.

@@

The distribution of codes in stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this number (267 and 146 stamps respectively). The workshops of the three \textit{conventus} show a similar pattern on the stamp frequency distribution with the exception of 2 large workshops in \textit{Hispalis} (La Catria and Arva). On these two workshops a comparatively large amount of stamp codes were found (29 different codes on each of them). According to previous studies, those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,arva_1997}.

It is worth mentioning that a majority of these stamps were collected during field surveys while only a hanfdul of workshops have been excavated; this may explain some of the differences in sample size as there could be intensity bias across the different sites~\citep{arva_1997}.

@@

The analysis of stamps in consumption areas also used the CEIPAC database and filters performed for the previous dataset. However, in this case there was a large amount of sites with only one or two stamps which may introduce noise in the similarity analysis. This is the reason why in this case an additional filter was performed and we only selected sites having at least five stamps.

The output was a dataset of 3386 stamps with similar sample size: 1765 from \textit{Britannia} and 1621 for the \textit{Germania} provinces. Both \textit{Germania} and \textit{Britannia} were analysed as frontiers and not as administrative Roman provinces; we included in the sample some centres that are actually located beyond the administrative boundaries but were considered part of the same border. Specifically, for the German \textit{limes} we included sites from the two provinces (Superior and Inferior) while the analysis of \textit{Britannia} extended to \textit{Caledonia}.

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
The 1765 stamps from \textit{Britannia} showed 986 unique codes from 46 sites. These consumption centres can be seen in Figure~\ref{britannia}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{britannia}
\caption{Sites in \textit{Britannia} where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. A majority of stamps have been found in garrisons related to Hadrian's and Antonine walls}
\label{britannia}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{germania}
\caption{Sites in Germania where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. Most sites are located along the German \textit{limes}}
\label{germania}
\end{figure}

@@

Similarity between amphora workshops and/or consumption sites was measured by quantifying a distance (also called dissimilarity) between the set of stamp codes found for each pair of sites (i.e. to what extent the stamp codes found on these sites were different). The chosen measure for this study was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of stamps while taking into account the frequency that each code is present on each site. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analysing their code.

@@

The result of the hierarchical clustering inferred from Morisita-Horn measures can be seen in Figure~\ref{dendro}. This visualisation suggests that most amphora workshops used a single stamp code which has not been found in any other site. Simultaneously the majority of stamp codes found on multiple sites are only present in two or three workshops. The combination of these two patterns generates a map of low similarity values for the entire sample.

@@

The lower similarity of stamp codes in \textit{Britannia} compared to \textit{Baetica} could be explained by a change of scale as the area under study is much bigger and the sites are more distant between them. The pattern can be more clearly observed for military sites which are often more well researched; in this case the ones with higher similarity tend to be geographically closer. To summarize, the same patterns than before can be observed on the dendrogram: 1) nearby sites tend to share more similar stamps and 2) a majority of stamp codes are only found in one site.

@@

Again, results in \textit{Germania} exhibit similar patterns with lower similarities. The only difference is that a higher concentration of sites sharing similar stamps was found in militarised areas near the \textit{limes}.

@@

The analysis of stamp similarity between workshops in \textit{Baetica} province suggests that a majority of stamp codes were only used in a single workshop. Beyond this clear pattern, it seems that the similarity of stamp codes is linked to proximity and the amphoric production of spatially close sites shared some stamp codes. Particularly relevant here is the fact that our results show that each \textit{conventus} area had similar stamps codes and they were not shared in other areas. This result could indicate that the workshops of the different \textit{conventus} used different stamp codes that were not shared between them. Within each conventus we can identify small groups of three/four workshops sharing the same amphoric stamps as seen in the dendrogam, but there are no stamp codes that may be found across several workshops. This pattern of similarity based on spatial closeness is closely aligned with previous studies that focused on morphometric distances between amphorae made on different workshops which also showed spatial correlation and large differences between \textit{conventus}~\citep{coto-sarmiento_identifying_2018}. Finally, while the role of the river was significant for the distribution of amphorae, river connectivity amongst workshops does not seem to show relevancy for the distribution or concentration of stamps in a specific river area.

On the other hand, our results do not seem to support the hypothesis that groups of workshops collectively used specific sets of stamp codes. There is higher similarities between nearby workshops but this could be simply explained by spatial proximity which would enable more frequent interactions. The opposite seems valid: each workshop used different stamp codes and only shared some of them with the closest ones.

This unique link between workshop and stamp code can be interpreted in different ways. First, it is possible that each workshop operated independently and did not normally collaborate with other workshops. Second, as discussedd stamp similarity in closer workshops could be linked to more intense interactions. Finally, the presence of different stamps in the same workshop could suggest some degree of internal organisation that would only affect nearby centres~\citep[104]{juanmorostesis}. For example, the stamping process could have allowed potters to sort out the different production batches and plan the shipments. Systematic procedures ike this one would probably be required to organise such a massive specialized production.

As we previously discussed some bias on the intensity of archaeological works and workshop identification may have an impact on this analysis. Further fieldwork could provide a clearer picture of the links between workshops if they allow to increase the sample size of stamp codes is increased.

@@

A combined exploration of the map and the dendrogram also suggests an interesting pattern: centres with higher similarity values are closer to the coast (North Sea and Celtic Sea). This may indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia}, since sites with higher degrees of similarity were strategic points near the sea.

@@

The areas in \textit{Britannia} with most intense military activity have some degree of overlap with civil areas. However, we do not detect this overlap for the the German \textit{limes} \citep{xanten2018}. This could be interpreted by parallel supply systems but it also could be caused by intensity bias as military centres have been more excavated than civil centres.

Finally, the results suggest that there is no clear organisational pattern linking production centres and/or consumption centres beyond local spatial proximity within regions. As a consequence, our analysis do not indicate the presence of specific Baetican amphoric workshop consistently supplying specific regions or even provinces.  This outcome suggests that stamp codes cannot be used to identify whether production centres had fixed clients and other proxies would be needed to identify this link in case it existed. Based on our analysis production centres could have distributed olive oil to both \textit{Britannia} and \textit{Germania} at a short-term basis of demand and without any long-term planning or centralised decision-making system.

>>>>

To achieve this goal, we analyse a dataset of stamps found in two types of sites: 1) workshops from \textit{Baetica} province and 2) consumption centres located in two provinces: \textit{Germania} (Inferior and Superior) and \textit{Britannia}. This dataset is used to detect links between specific sites that may share a large percentage of common stamps. The analysis is performed using quantitative methods borrowed from Ecology that allows us to group sites based on the number of stamps shared between them.

@@

The finding of identical amphoric stamps in different provinces has allowed us to explore patterns of olive oil distribution within the Empire. Despite this archaeological evidence, there is still no consensus on the structure that generated this pattern; how were the economic relations between production and consumption centres? Did they follow a centralised process to distribute olive oil in those provinces or each workshop organised its own trade networks?
%repite aims dos veces  
This paper aims to explore these questions regarding the olive oil market by identifying links between production centres and consumption provinces using similarity indexes for amphoric stamps. In this way, our work aims to detect regional patterns for a specific commercial product (i.e. olive oil) which was shipped between distant provinces~\citep{isaksen_network_2006}.

This relationship between the production centre (i.e. \textit{Baetica}) and consumption sites have been studied here for two case studies: \textit{Britannia} and \textit{Germania} (see Figure~\ref{general}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{general_map}
\caption{Sites analysed in this paper. The colour defines the three different regions under study (\textit{Baetica} (yellow) as producer; \textit{Germania} (blue) and \textit{Britannia} (green) as consumers) while the size of each dot is proportional to the number of Dressel 20 amphoric stamps found in each site}

@@

In \textit{Baetica} province we want to identify links between workshops based on the similarity of their stamps. In the consumption Roman provinces, our aim is to detect groups of nearby sites sharing stamps and to what extent stamps are clustered based on provinces. This economic connection could be identified with two ideas: a) nearby sites have more similar amphoric stamps and b) sets of stamps can only be found in specific provinces because they were shipped together.
%find in or find on

@@

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by identifying which centres have more similar stamps and then confront this measure against their proximity. To do this, a population-based approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. This quantitative method provides a way to measure the similarity between centre based on the amphoric stamps found in each of them. This ecological approach is based on three steps: a) to quantify stamp similarity between sites, b) to group sites based on this similarity and c) to find correlation patterns between stamp similarity and proximity. The rationale behind the method is as follows: if trade structure was centrally planned then we should be able to establish links between workshops and consumption sites based on the similarity of its stamps; if this is not the case then results would suggest that workshops worked independently and agreements were made at an individual level.

The paper addresses these questions, the analysis and results as follows: next sections introduce the historical context. Section five defines the dataset and the methods used for the analysis. Section six presents the results and section seven discusses them; the paper finishes with some concluding remarks on the research performed.

@@

The province of \textit{Baetica} was one of the most suitable regions to tackle the challenge of the huge demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive specialised infrastructure of olive oil production was gradually deployed. This production of olive oil had the largest peak from the first to the third century AD~\citep{remesal_concierto}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{baetica}
\caption{Distribution of stamps found in Dressel 20 workshops along the Guadalquivir riverbank. The size of each point depends on the number of stamps found in each site}

@@

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were typically marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure} 
%found in find in
Most stamps seemed to have been used across a long activity period which can be difficult to date with the required level of accuracy. They have been frequently dated using as reference consular dates found in nearby \textit{tituli picti} recovered from \textit{Monte Testaccio}~\citep{Testaccio1, berni_millet_epigrafianforica_2008}.
The chronological information of \textit{tituli picti} is very useful, but the small amount of dates recovered and the fact that a vast majority of them come from the same site poses some challenges to this precise dating. In addition, the fact that \textit{Monte Testaccio} was a dump suggests that the disposition of amphorae vessels may have been altered at a later phase due to changes in the long-term dynamics of creation of this artificial mound.

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found in handles but they could also be imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps was not defined by a single criteria, but they often displayed a code composed by letters and additional symbols. One of the most common forms is a code of three letters showing a \textit{Tria Nomina}~\citep{berni_millet_amphora_1996}; in this case the three letters are the initials of a Roman name. In some cases the full name may appear, while the meaning of other stamps is difficult to interpret~\citep{rodriguez_baetican_1998}.

@@

The expansion of the Roman Empire to increasingly distant regions required an effort to intensify trade activity through Mediterranean and Atlantic routes. These needs effectively led to a gradual change in the economic and social structures as the trade networks that supplied the provinces and particularly the Roman army became more complex. An interesting example of this system is the figure of the \textit{praefectura annonae} which was created during Augustus's administration to organise wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced in the region as the landscape and climate of the British Isles was not suitable for the growth of olive trees~\citep[161]{monfort_britanniaen_1998}. Archaeological records suggest that this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century AD onwards \citep{peacock_amphorae_1991,carreras_britannia_1998}. At this moment, we detect an increase in the olive oil exports linked to the arrival of legions caused by different military campaigns~\citep[161]{monfort_britanniaen_1998}.

The increase in Dressel 20 exports created an important commercial network for exchanges. The demand for olive oil after the Roman conquest required a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply. This economic relationship will be particularly intense for sites close to Hadrian Wall's garrisons. Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic locations~\citep{carreras_atlantic_2012}. This presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,funari_corpus_1996,carreras_britannia_1998,carreras_abastecimiento_2003}. This intense consumption of olive oil also indicates the existence of provincial structures created to organise the supply of olive oil across military camps. No written records detailing how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep{funari_economic_2005,orengo_seeds_2016}.

The consumption of Baetican olive oil would experience a progressive slowdown from the third century AD onwards. This date matches a change in market strategy in the Empire. This gradual decrease in \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by other types of amphorae from other places and particularly the African provinces~\citep{rodriguez1991aceite,millet_anforas_1998}.

@@

The distribution of Dressel 20 amphorae in the provinces of \textit{Germania Inferior} and \textit{Germania Superior} is not as well interpreted as the rest of the provinces discussed in this work~\citep[293]{remesal_baetica_2002}. 
%it is clear no se si ponerlo
It is clear that there was a massive import of Dressel 20 in both provinces as is observed in archaeological excavations across the German \textit{limes}. However, it is still unknown to what extent trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}.

In any case, the introduction of olive oil in \textit{Germania} shows similar patterns to the one discussed for\textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region, while a majority of recovered Dressel 20 amphorae are located in the military sites built over the German \textit{limes} \citep{remesal_germaniaengl_2002}. The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a major concentration at the German \textit{limes}. It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons~\citep[125]{remesal_concierto}. However, this hypothesis is hard to assess given the current lack of written evidence clarifying the meaning of the amphoric stamps.

Some studies assumed that Roman naval technology was not suited for oceanic transport and for this reason defined riverine shipping across Rh\^one and Rhine rivers as the basic supply route for the German \textit{limes} \citep{marliere2001tonneau}. However, more recent large-scale computational analysis, as well as new excavations on the Iberian Atlantic coast suggest that the Atlantic route, which was far more efficient, may have played a major role in the supply of \textit{Germania} and \textit{Britannia}~\citep{rubio-campillo_ecology_2018}.

@@

The goal of this study is to explore these economic and long-range trade dynamics using amphoric stamps as proxies for common supply networks. We are especially interested in identifying links between Baetican production centres and consumption centres in the different \textit{limes}.

For this analysis, we used data collected in the CEIPAC database of amphoric epigraphy~\citep{remesal_centro_2015}. The CEIPAC dataset contains over 50.000 epigraphic records found on different types of amphorae (the database is openly accessible from \url{http://romanopendata.eu}). A large percentage of these records come from \textit{Monte Testaccio}, but the rest is distributed mainly across the Western provinces of the Empire. It must be mentioned that the database does not provide temporal information due to the reasons previously discussed, and for this reason, this work can only analyse the spatial distributions within a coarse chronology of three centuries (1st-3rd AD).

@@

It is worth noting that some workshops were coded under the same geographical coordinates. For instance, the stamps of two nearby amphora workshops such as Tesorillo de Do\~na Mencia and Do\~na Mencia were catalogued under different names while sharing the same geographical coordinates. The interpretation of our analysis considered this bias in the archaeological dataset which may have a minor effect on the final results.

The first step of this analysis was to compute the frequency distribution of stamp codes and analyse it using Exploratory Data Analysis (EDA). This would allow us to study the distribution of the stamps across centres as can be seen in Figure~\ref{stamps}. The distribution is widely diverse with most workshops having only one stamp, while a few workshops contained over 100 stamps. Particularly, one workshop (La Catria) concentrated a large percentage of the amphoric stamps with a total number of 228 stamps.

This uneven distribution of stamps is typical for amphora production as it is observed that a major concentration on the number of stamps is focused on a few workshops~\citep{bayesian_2018,coto-sarmiento_identifying_2018}.

@@

The distribution of codes in stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this number (267 and 146 stamps respectively). The workshops of the three \textit{conventus} show a similar pattern on the stamp frequency distribution with the exception of two large workshops in \textit{Hispalis} (La Catria and Arva). In these two workshops a comparatively large amount of stamp codes were found (29 different codes on each of them). According to previous studies, those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,arva_1997}.

It is worth mentioning that a majority of these stamps were collected during field surveys while only a handful of workshops have been excavated; this may explain some of the differences in sample size as there could be intensity bias across the different sites~\citep{arva_1997}.

@@

The analysis of stamps in consumption areas used also the CEIPAC database and filters performed for the previous dataset. However, in this case there was a large number of sites with only one or two stamps which may introduce noise in the similarity analysis. This is the reason why in this case an additional filter was performed and we only selected sites having at least five stamps.

The output was a dataset of 3386 stamps with similar sample size: 1765 from \textit{Britannia} and 1621 for the \textit{Germania} provinces. Both \textit{Germania} and \textit{Britannia} were analysed as frontiers and not as administrative Roman provinces; we included in the sample some centres that are located beyond the administrative boundaries but were considered part of the same border. Specifically, for the German \textit{limes} we included sites from the two provinces (Superior and Inferior) while the analysis of \textit{Britannia} extended to \textit{Caledonia}.

The 1765 stamps from \textit{Britannia} showed 986 unique codes from 46 sites. These consumption centres can be seen in Figure~\ref{britannia}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{britannia}
\caption{Sites in \textit{Britannia} where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found in the site. A majority of stamps have been found in garrisons related to Hadrian's and Antonine walls}
\label{britannia}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{germania}
\caption{Sites in \textit{Germania} where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found in the site. Most sites are located along the German \textit{limes}}
\label{germania}
\end{figure}

@@

Similarity between amphora workshops and/or consumption sites was measured by quantifying a distance (also called dissimilarity) between the set of codes of stamps found for each pair of sites (i.e. to what extent the stamp codes detected in these sites were different). The chosen measure for this study was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of stamps while taking into account the frequency that each code is present on each site. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analysing their code.

@@

The result of the hierarchical clustering inferred from Morisita-Horn measures can be seen in Figure~\ref{dendro}. This visualisation suggests that most amphora workshops used a single stamp code which has not been found in any other site. Simultaneously the majority of stamp codes found in multiple sites are only present in two or three workshops. The combination of these two patterns generates a map of low similarity values for the entire sample.

@@

The lower similarity of stamp codes in \textit{Britannia} compared to \textit{Baetica} could be explained by a change of scale as the area under study is much bigger and the sites are more distant between them. The pattern can be more clearly observed for military sites which are often more well researched; in this case the ones with higher similarity tend to be geographically closer. To summarise, the same patterns than before can be observed in the dendrogram: 1) nearby sites tend to share more similar stamps and 2) a majority of stamp codes are only found in one site.

@@

Additionally, results in \textit{Germania} exhibit similar patterns with lower similarities. The only difference is that a higher concentration of sites sharing similar stamps was found in militarised areas near the \textit{limes}.

@@

The analysis of stamp similarity between workshops in \textit{Baetica} province suggests that a majority of stamp codes were only used in a single workshop. Beyond this clear pattern, it seems that the similarity of stamp codes is linked to proximity and the amphoric production of spatially close sites shared some stamp codes. Particularly relevant here is the fact that our results show that each \textit{conventus} area had similar stamps codes and they were not shared in other areas. This result could indicate that the workshops of the different \textit{conventus} used different stamp codes that were not shared between them. Within each \textit{conventus} we can identify small groups of three/four workshops sharing the same amphoric stamps as seen in the dendrogram, but there are no stamp codes that may be found across several workshops.

This pattern of similarity based on spatial closeness is closely aligned with previous studies that focused on morphometric distances between amphorae made on different workshops which also showed spatial correlation and large differences between \textit{conventus}~\citep{coto-sarmiento_identifying_2018}. Finally, while the role of the river was significant for the distribution of amphorae, river connectivity amongst workshops does not seem to show relevancy for the distribution or concentration of stamps in a specific river area.

On the other hand, our results do not seem to support the hypothesis that groups of workshops collectively used specific sets of stamp codes. There are higher similarities between nearby workshops but this could be simply explained by spatial proximity which would enable more frequent interactions. The opposite seems valid: each workshop used different stamp codes and only shared some of them with the closest ones.

This unique link between workshop and stamp code can be interpreted in different ways. First, it is possible that each workshop operated independently and did not normally collaborate with other workshops. Second, as discussed stamp similarity in closer workshops could be linked to more intense interactions. Finally, the presence of different stamps in the same workshop could suggest some degree of internal organisation that would only affect nearby centres~\citep[104]{juanmorostesis}. For example, the stamping process could have allowed potters to sort out the different production batches and plan the shipments. Systematic procedures like this one would probably be required to organise and control such a massive specialised production.

As we previously discussed some bias on the intensity of archaeological works and workshop identification may have an impact on this analysis. Further fieldwork could provide a clearer picture of the links between workshops if they allow increasing the sample size of stamp codes is increased.
%mirar el increased que creo que está mal puesto

@@

A combined exploration of the map and the dendrogram also suggests an interesting pattern: centres with higher similarity values are closer to the coast (North Sea and Celtic Sea). This may indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia} since sites with higher degrees of similarity were strategic points near the sea.

@@

The areas in \textit{Britannia} with most intense military activity have some degree of overlap with civil areas. However, we do not detect this overlap for the German \textit{limes} \citep{xanten2018}. This could be interpreted by parallel supply systems but it also could be caused by intensity bias as military centres have been more excavated than civil centres.
%civil centres no sé si está bien
Finally, the results suggest that there is no clear organisational pattern linking production centres and/or consumption centres beyond local spatial proximity within regions. As a consequence, our analysis does not indicate the presence of specific Baetican amphoric workshop consistently supplying specific regions or even provinces. This outcome suggests that stamp codes cannot be used to identify whether production centres had fixed clients and other proxies would be needed to identify this link in case it existed. Based on our analysis production centres could have distributed olive oil to both \textit{Britannia} and \textit{Germania} at a short-term basis of demand and without any long-term planning or centralised decision-making system.
